\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-12-14 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated pairwise comparisons, simultaneous CI for smooths

#creates a dataframe using the length of the covariates for the GAM
gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1),
                         subject=factor(rep(1:10)))

@@

#adds the predictions to the grid and creates a confidence interval for GAM
gam_predict<-gam_predict%>%
    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=al,
              data=gam_predict,
            show.legend=FALSE,
            inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

@@

m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
m_predict<-m_predict%>%
    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=al,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

@@

mult_plot<-f2+f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-7,75)) &
  theme(
     text=element_text(size=18)
     )

@@

```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
##Pairwise comparisons

pdat <- expand.grid(Day = seq(0, 10, length = 400),
                    Group = c('Control', 'Treatment'))

#this function takes the model, grid and groups to be compared using the lpmatrix

smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    #Find columns in xp where the name contains "Control" and "Treatment"
    col1 <- grepl(g1, colnames(xp))
    col2 <- grepl(g2, colnames(xp))
    #Find rows in xp that correspond to each treatment
    row1 <- with(newdata, Group == g1)
    row2 <- with(newdata, Group == g2)
    ## difference rows of xp for data from comparison
    X <- xp[row1, ] - xp[row2, ]
   ## zero out cols of X not involved in the comparison
    X[, ! (col1 | col2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(g1, g2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')

comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp1)) %>%
  mutate(interval=case_when(
    upper>0 & lower<0~"no-diff",
    upper<0~"less",
    lower>0~"greater"
  ))

#function to obtain values for the shading regions
pairwise_limits<-function(dataframe){
    #extract values where the lower limit of the ribbon is greater than zero
    #this is the region where the control group effect is greater
    v1<-dataframe%>%
        filter(lower>0)%>%
        select(Day)
    #get day  initial value
    init1=v1$Day[[1]]
    #get day final value
    final1=v1$Day[[nrow(v1)]]

    #extract values where the value of the upper limit of the ribbon is lower than zero
    #this corresponds to the region where the treatment group effect is greater
    v2<-comp_StO2_full%>%
        filter(upper<0)%>%
        select(Day)

    init2=v2$Day[[1]]
    final2=v2$Day[[nrow(v2)]]
    #store values
    my_list<-list(init1=init1,
                  final1=final1,
                  init2=init2,
                  final2=final2)
return(my_list)
}

my_list<-pairwise_limits(comp_StO2_full)
rib_col<-'#8D7D82' #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E' #color for ribbon treatment region

c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
    annotate("rect",
                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
                fill=control_rib,
                alpha = 0.5,
                ) +
  annotate("text",
             x=1.5,
             y=-18,
             label="Control>Treatment",
           size=8,
           angle=90
           )+
    annotate("rect",
             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5,
    ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=8,
           angle=90
           )+
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill=rib_col) +
    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
    geom_line(color='black',size=1) +

    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

###for missing data
comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', 
         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
                              )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme_classic()+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )

my_list<-pairwise_limits(comp_StO2_missing)

c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    annotate("rect",
             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
             fill=control_rib,
             alpha = 0.5
             ) +
  annotate("text",
             x=1.5,
             y=-18,
             label="Control>Treatment",
           size=8,
           angle=90
           )+
    annotate("rect",
             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
             fill= treat_rib,
             alpha = 0.5,
    ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=8,
           angle=90
           )+
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill=rib_col) +
    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

pair_comp+plot_annotation(tag_levels='A')&
  theme(
     text=element_text(size=18)
     )

>>>>

## point-wise interval
ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
## simultaneous interval
si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)

# mean shift for Treatment group
const <- coef(gam_02)[2]

#pointwise confidence interval
ci <- ci %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

#simultaneous interval
si <- si %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

@@

@@

#plot smooths and pointwise and simulatenous confidence intervals for GAM
f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.4,
show.legend=FALSE)+
    geom_point(data=dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group), 
                        size=1.5,
                        alpha=0.6, 
               inherit.aes = FALSE,
               show.legend = FALSE)+
    labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

@@

ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
## simultaneous interval
si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)

# mean shift for group 2
const <- coef(mod_m1)[2]

ci <- ci %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

si <- si %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.4,
show.legend=FALSE)+
    geom_point(data=dat_missing, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group), 
                            size=1.5,
                        alpha=0.6, 
               inherit.aes = FALSE)+
    labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22))+
        thm1

@@

mult_plot<-f2+f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-20,75)) &
  theme(
     text=element_text(size=18)
     )

@@

```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}

##Pairwise comparisons

calc_difference_R <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
  ## make sure f1 and f2 are characters
  f1 <-  as.character(f1)
  f2 <-  as.character(f2)
  cnames <- colnames(Xp)
  ## columns of Xp associated with pair of smooths
  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
  ## rows of Xp associated with pair of smooths
  r1 <- data[[by_var]] == f1
  r2 <- data[[by_var]] == f2

  ## difference rows of Xp for pair of smooths
  X <- Xp[r1, ] - Xp[r2, ]

  ## zero the cols related to other splines
 # X[, ! (c1 | c2)] <- 0

  ## zero out the parametric cols
  #X[, !grepl('^s\\(', cnames)] <- 0

  ## compute difference
  sm_diff <- drop(X %*% coefs)
  se <- sqrt(rowSums((X %*% V) * X))
  nr <- NROW(X)

  ## Calculate posterior simulation for smooths
  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
  rownames(coefs_sim) <- rownames(V)
  simDev <- X %*% coefs_sim
  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
  masd <- apply(absDev, 2, max)
  crit_s <- quantile(masd, prob = 0.95, type = 8)

  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
              level_1 = rep(f1, nr),
              level_2 = rep(f2, nr),
              diff = sm_diff, se = se,
              lower_s = sm_diff - crit_s * se,
              upper_s = sm_diff + crit_s*se)

  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
  ## Only need rows associated with one of the levels
  out <- bind_cols(out, data[r1, smooth_var])

  out
}

#does both ci and si
difference_smooths_R <- function(model,
                                 smooth,
                                 n = 100,
                                 ci_level = 0.95,
                                 newdata = NULL,
                                 partial_match = TRUE,
                                 unconditional = FALSE,
                                 frequentist = FALSE,
                                 nrep = 10000,
                                 include_means = TRUE,
                                 ...) {
  if (missing(smooth)) {
    stop("Must specify a smooth to difference via 'smooth'.")
  }

  # smooths in model
  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
  # select smooths
  select <-
    gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                       partial_match = partial_match)#,
  # model_name = expr_label(substitute(object)))
  sm_ids <- which(select)
  smooths <- gratia::get_smooths_by_id(model, sm_ids)
  sm_data <- map(sm_ids, gratia:::smooth_data,
                 model = model, n = n, include_all = TRUE)
  sm_data <- bind_rows(sm_data)
  by_var <- by_variable(smooths[[1L]])
  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
                                   stringsAsFactor = FALSE))
  names(pairs) <- paste0("f", 1:2)

  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
  V <- gratia:::get_vcov(model, unconditional = unconditional,
                         frequentist = frequentist)
  coefs <- coef(model)

  out <- pmap(pairs, calc_difference_R, smooth = smooth, by_var = by_var,
              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
              coefs = coefs, nrep = nrep)
  out <- bind_rows(out)
  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)

  out <- add_column(out,
                    lower = out$diff - (crit * out$se),
                    upper = out$diff + (crit * out$se),
                    .after = 6L)
  out
}

#compute difference between smooths and calculate confidence interval: complete data
diff_complete <- difference_smooths_R(gam_02, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)

#function to obtain values for the shading regions
pairwise_limits<-function(dataframe){
    #extract values where the lower limit of the ribbon is greater than zero
    #this is the region where the control group effect is greater

    v1<-dataframe%>%
        filter(lower_s>0)%>%
        select(Day)
    #get day  initial value
    init1=v1$Day[[1]]
    #get day final value
    final1=v1$Day[[nrow(v1)]]
    #extract values where the value of the upper limit of the ribbon is lower than zero
    #this corresponds to the region where the treatment group effect is greater
    v2<-dataframe%>%
        filter(upper_s<0)%>%
        select(Day)

    init2=v2$Day[[1]]
    final2=v2$Day[[nrow(v2)]]
    #store values
    my_list<-list(init1=init1,
                  final1=final1,
                  init2=init2,
                  final2=final2)
    return(my_list)

}

my_list<-pairwise_limits(diff_complete)
rib_col<-'#8D7D82' #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E' #color for ribbon treatment region

c1 <- ggplot() +
  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
    annotate("rect",
                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
                fill=control_rib,
                alpha = 0.5,
                )+
    annotate("text",
             x=1.5,
             y=-18,
             label="Control>Treatment",
           size=8,
           angle=90
           )+
    annotate("rect",
             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5
             ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=8,
           angle=90
           )+
  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
    axis.text=element_text(size=22))+
    theme_classic()

###compute difference in smooths for missing data
diff_missing <- difference_smooths_R(mod_m1, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)

v2<- diff_missing %>%
        filter(upper_s<0)%>%
        select(Day)
init2=v2$Day[[1]]
final2=v2$Day[[nrow(v2)]]

c2 <- ggplot() +
  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
    annotate("rect",
             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5,
    ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=8,
           angle=90
           )+
  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
    axis.text=element_text(size=22))+
    theme_classic()

@@

pair_comp+plot_annotation(tag_levels='A')&
    theme(
     text=element_text(size=18)
     )
```
